Extracting Function Signatures from Program Traces

### 2020-05-14 paper/main.tex
fixes

In R, function can be called with named parameters, R support variable
argument lists, and arguments can have default values. Putting all of these
together consider the following declaration:

@@

R has one builtin notion of type that can be queried by the \k{typeof}
function. Over the years, programmers have found the need for a richer type
structure and have added {\it attributes}. The best way to think of attributes is
as an optional map from name to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.

@@

Figure~\ref{types} lists all of the builtin types that are provided by the
language. They are the possible return values of \k{typeof}. There is no
intrinsic notion of subtyping in R. But, in many context a \k{logical} will
convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
off conversion can occur in corner cases, such as \k{1<"2"} holds and
\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
converted to an integer. R does not distinguish between scalars and vectors
(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
distinguished missing value denoted by \code{NA}. The default type of
\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.

@@

The addition of attributes lets programmers extend the set of types by
tagging data with user-defined attributes. For example, one could define a
vector of four values, \code{x<-c(1,2,3,4)} and then attach the attribute
\k{dim} with a pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From
that point, arithmetic functions will treat \k{x} as a 2x2 matrix. Another
attribute that can be set is the \k{class}.  This attribute can be bound to
a list of class names. For instance, \code{class(x)<-"human"}, set the class
of \k{x} to be \k{human}.  Attributes are thus used for object-oriented
programming. The S3 object system support single dispatch on the class of
the first argument of a function, whereas the S4 object system allows
multiple dispatch (on all arguments). Some of the most widely used data
type, such as data frames, leverage attributes. A data frame, for instance,
is a list of vectors with a class and a column name attribute.

@@

On the subject of matrices, they represent another layer of complexity on top of vectorized primitives.
Internally, matrices are simply vectors with class ``matrix'' and a \code{dims} attribute indicating matrix dimensions, and many functions automatically coerce vectors to matrices when appropriate (while not codified in the language semantics, many internal functions perform the coercion).
Relying on automatic coercion is unwise, and many functions check to ensure that their arguments are matrices, particularly those in packages which rely on matrices.
For instance, consider the code in Figure~\ref{fig:matrix-example}.

@@

The \code{rowWeightedMeans} function calculated the weighted means of rows or columns of a matrix \code{x}.
This function exhibits good practice in type checking arguments, rightly producing a message to the user indicating that an explicit matrix should be passed.

To capture cases like this, we include a \code{class<matrix>} type in our type annotation language.
If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically.
That said, we don't intend to subsume all argument validation code, discussed next.

@@

Given the discussion of scalars, vectors, and matrices, it might be tempting to consider factoring dimensionality into types.
The primary drawback of this would be rendering our type annotation language into a language of dependent types, which we believe too complex to retrofit onto a dynamic data science language such as R.  
This does mean our types cannot subsume all validation code, such as some later code from the aforementioned \code{rowWeightedMeans} function, shown in Figure~\ref{fig:general-validation-example}.

@@

Here, the length of the weights vector \code{w} is being checked for equality with the number of columns in the matrix \code{x}. 
Ultimately, we decided to leave these kinds of checks up to the programmer.
\AT{I feel like we might want a better argument to not make our type language dependent (better than lol pls no dependent types), but I could be wrong.
Maybe reviewers would be ok with a simple argument.}

@@

We are faced with several possible designs for list types.  
Chief concerns include the heterogeneity of list elements, and the presence of a \code{names} attribute, allowing list elements to be accessed in a struct-like manner.  
\AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.  
The next sections highlight some important design decisions for our list types.

@@

@@

%
%
\subsubsection{Forgoing Tuples, or Short, Constant Length Lists}

We briefly entertained the possibility of including a \code{tuple} type, which would have looked like e.g. \code{tuple<int, int>} for a list of two integers.  
When configured to detect tuples, our analysis did indeed pick up \AT{a number}, but post-processing and simplification revealed that in \AT{many} cases, they would be subsumed by co-occurring \code{list} types.
In other words, more often than not it just so happened that functions were being called with short lists.  
Of the cases that tuples were not subsumed by lists, \AT{often} there were \AT{tuples of different lengths}, implying that the shape of the argument was not essential.

\AT{I feel like this is the least important of the design decisions we made, so I'm leaving it for the end, we might not even want it.}

@@

Another pattern that we observed was the use of consistently-named lists.
In R, ``fields'' of named lists can be accessed using the \code{$} operator, as illustrated in the code snippet in Figure~\ref{fig:struct-ex}.

@@

\begin{lstlisting}
calculate.neuralnet <- function (data, model.list, hidden, rep,
            startweights, exclude, constant.weights) {
    result <- generate.startweights(model.list, hidden, startweights,
                                    rep, exclude, constant.weights)
    weights <- result$weights
    exclude <- result$exclude
    ...
\end{lstlisting}

\caption{Struct example from the \code{neuralnet} package.}
\label{fig:struct-ex}
\end{center}
\end{figure}

>>>>

In R, functions can be called with named parameters, arguments can have default values, and R also support variable argument lists. 
To illustrate all of this in a single example, consider:

@@

R has one builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{types} lists all of the builtin types that are provided by the
language. They are the possible return values of \k{typeof}. There is no
intrinsic notion of subtyping in R. But, in many context a \k{logical} will
convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
off conversion can occur in corner cases, such as \k{1<"2"} holds and
\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
converted to an integer. R does not distinguish between scalars and vectors
(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
distinguished missing value denoted by \code{NA}. The default type of
\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.

@@

@@

Over the years, programmers have found the need for a richer type
structure and have added {\it attributes}. The best way to think of attributes is
as an optional map from name to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.
The addition of attributes lets programmers extend the set of types by
tagging data with user-defined attributes. For example, one could define a
vector of four values, \code{x<-c(1,2,3,4)} and then attach the attribute
\k{dim} with a pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From
that point, arithmetic functions will treat \k{x} as a 2x2 matrix. Another
attribute that can be set is the \k{class}.  This attribute can be bound to
a list of class names. For instance, \code{class(x)<-"human"}, set the class
of \k{x} to be \k{human}.  Attributes are thus used for object-oriented
programming. The S3 object system support single dispatch on the class of
the first argument of a function, whereas the S4 object system allows
multiple dispatch (on all arguments). Some of the most widely used data
type, such as data frames, leverage attributes. A data frame, for instance,
is a list of vectors with a class and a column name attribute.

@@

Matrices represent another layer of complexity on top of vectorized primitives.
Internally, matrices are simply vectors with class ``matrix'' and a \code{dims} attribute indicating matrix dimensions, and while not codified in the language semantics, many internal functions will coerce vectors to matrices automatically.
Relying on automatic coercion is unwise, and many functions check to ensure that their arguments are matrices.
For instance, consider the code in Figure~\ref{fig:matrix-example}.

@@

The \code{rowWeightedMeans} function calculates the weighted means of rows or columns of a matrix \code{x}.
This function exhibits good practice in type checking arguments, rightly producing a message to the user indicating that an explicit matrix should be passed.

To capture cases like this, we include a \code{class<matrix>} type in our type annotation language.
If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically, notifying the users passing ill-typed arguments.

@@

Thus far we've seen instances where we introduce types that subsume existing argument validation code, but our intention is not to subsume {\it all} such code.
As an example of something we {\it do not} want to do, consider the code in Figure~\ref{fig:general-validation-example}.

@@

This code is from later in the same \code{rowWeightedMeans} function discussed previously.
Here, the length of the weights vector \code{w} is being checked for equality with the number of columns in the matrix \code{x}. 
Specifying this relationship as a type amounts to {\it dependent typing}, which we believe too complex to retrofit onto a dynamic data science language such as R.  
Ultimately, we decided to leave these kinds of checks up to the programmer.
\AT{Can we say more here?}

@@

We are faced with several possible designs for list types.  
Chief concerns include the heterogeneity of list elements and the presence list element names, which allow list elements to be accessed in a struct-like manner.  
% \AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.  
The next sections highlight some important design decisions for our list types.

@@

\AT{Worth a code example?}

@@

%
%
% \subsubsection{Forgoing Tuples, or Short, Constant Length Lists}

% We briefly entertained the possibility of including a \code{tuple} type, which would have looked like e.g. \code{tuple<int, int>} for a list of two integers.  
% When configured to detect tuples, our analysis did indeed pick up \AT{a number}, but post-processing and simplification revealed that in \AT{many} cases, they would be subsumed by co-occurring \code{list} types.
% In other words, more often than not it just so happened that functions were being called with short lists.  
% Of the cases that tuples were not subsumed by lists, \AT{often} there were \AT{tuples of different lengths}, implying that the shape of the argument was not essential.

% \AT{I feel like this is the least important of the design decisions we made, so I'm leaving it for the end, we might not even want it.}

@@

We observed the use of consistently-named lists in \AT{a number of functions}.
In R, ``fields'' of named lists can be accessed using the \code{$} operator, as illustrated in the code snippet in Figure~\ref{fig:struct-ex}.

@@

\begin{lstlisting}
agricolae::cv.model <- function(x) {
  suma2 <- sum(x$residual^2)
  gl <- x$df.residual
  promedio <- mean(x$fitted.values)
  return(sqrt(suma2/gl)*100/promedio)
}

cars_model <- lm(dist ~ speed, data = cars)
cv.model(cars_model)
\end{lstlisting}

\caption{Struct example from the \code{agricolae} package.}
\label{fig:struct-ex}
\end{center}
\end{figure}

In the code snippet, the function \code{cv.model} takes an argument \code{x}, which we observed to always be a {\it linear model}, which is part of R's built-in statistics functionality.
Internally, linear models are represented as lists with named elements, and one may access these named elements using the \code{$} syntax, as seen on lines 2-4 in the snippet.
